---
title: "Exercise 1 --- Basic Maps and Graphs in R"
subtitle: "BSTA 374 · 12 points · Due Thursday, September 10, 2026, 11:59 pm"
author: "Eric M. Delmelle · Lehigh University"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    toc-location: left
    toc-title: "On this page"
    number-sections: false
    code-copy: true
    highlight-style: github
    embed-resources: true
  pdf:
    papersize: letter
    geometry:
      - margin=1in
    toc: false
    colorlinks: true
    code-overflow: wrap
prefer-html: true
execute:
  eval: false
  echo: true
---

## What this exercise is

Your first graded work in this course. It covers the first two weeks: reading
spatial data with `sf`, understanding what a coordinate reference system is and
moving between them, and producing honest choropleth and point maps with both
`ggplot2` and `tmap`.

**No data to download.** Everything runs on the North Carolina SIDS data that
ships inside the `sf` package, so your analysis will reproduce on my machine
exactly as it does on yours --- which is the point.

Each part ends with **starting code**. Copy it, fill in the blanks marked
`______`, and adapt it. It is scaffolding, not a solution: the mechanical lines
are given so that your time goes into the parts that require judgement.

::: {.callout-important}
## Submission
Submit **one file** to Course Site:

- `LASTNAME_FI_Exercise1.qmd` --- the source

Example: `DELMELLE_EM_Exercise1.qmd`. Put your name in the document header.

**Due Thursday, September 10, 2026 at 11:59 pm.** Late work loses 20% per day
unless you arrange something with me in advance.
:::

::: {.callout-warning}
## Reproducibility is graded
Before you submit: restart R (`Ctrl/Cmd + Shift + F10`) and render the document
from a clean session. If it does not run start to finish, it is incomplete ---
regardless of how good the maps are. This is the habit that matters most, and
we start enforcing it now.
:::

```{r}
#| label: setup

library(sf)
library(dplyr)
library(ggplot2)
library(tmap)
# add anything else you use -- and load it here, not halfway down

nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
st_crs(nc)   
```

---

## Part A --- Read and describe the data (2 points)

Answer each in **one or two sentences of prose**, outside the code chunks, and
show the code you used to find out.

1. How many features does `nc` contain, and what does one row represent?
nc has 15 columns and 100 counties/features, and one row represents the data for each county in North Carolina. 

2. What geometry type is stored? Why is it that type rather than `POLYGON`?
   *(Look at the coast.)*
The geometry type that is stored to create the shape of the NC counties is a multipolygon as some counties are comprised of multiple polygons/shapes, such as on the coast. 
   
3. What coordinate reference system is the data in? Give its name, say whether it
   is geographic or projected, and say how you found out.
I used st_crs(nc) and saw that the data uses NAD27/4267. It is a geographic CRS because the output of the st_crs() function says that it uses latitude and longitude, as mentioned in Ch 2 of the reading. 
   
4. In your own words, what do `BIR74` and `SID74` measure? Why does having both
   matter more than having either one alone?
BIR74 is the births in each county, while SID74 is the number/count of sudden infant deaths in that county in that time period. Without BIR74 you cannot account for population size/number of live births through calculating a rate for SID74. 

### Starting code

```{r}
#| label: a-starter

# A1 -- how big is it, and what are the columns?
nrow(nc)
names(nc)
nc                       # the header block above the table says a lot

# A2 -- one geometry type per feature; count them
table(st_geometry_type(nc))

# A3 -- three ways of asking about the CRS
st_crs(nc)               # the full definition
st_crs(nc)$input         # the short name
st_is_longlat(nc)        # TRUE = geographic (degrees), FALSE = projected (metres)

# A4 -- the help page documents every column
?nc
```

---

## Part B --- Coordinate reference systems (2 points)

1. Transform `nc` to **EPSG:32119** (NAD83 / North Carolina State Plane, in
   metres) and store it as a **new object**. Do not overwrite `nc` --- you need
   both.
   
2. Plot the counties **before and after**, near enough to each other that a reader can compare them. One or two sentences on what visibly changed.
  On the right side (the projected version with NAD83), North Carolina's northern border appears to be more curved, possibly better accounting for the shape of the Earth but I am not 100% sure. On the left side, the unprojected geographic version is less curved. 
   
3. Compute the area of every county in **square kilometres**, using the projected
   version. Report the five largest.
The five largest counties in km ^2 are calculated as Columbus, Robeson, Sampson, Bladen, and Wake 
   
4. Now compute areas from the *unprojected* object as well. Compare the two sets
   of numbers and explain, in two or three sentences, what `sf` did and why the
   projected version is the one you would report in a paper.
The values for the unprojected and projected maps are quite similar, however the values for the projected nc seem to be slightly larger for each county. This is because sf calculates area based on the specific CRS that was given. In the case of our projected vs. non-projected nc's, I would choose nc_proj as it uses North Carolina's specific CRS, meaning I would think it would be more accurate. 


::: {.callout-note}
Step 4 will not give you an error, and the two sets of numbers will be close but
not identical. That is the interesting part --- the question is which *surface*
each answer refers to.
:::

### Starting code

```{r}
#| label: b-starter

# B1
nc_proj <- st_transform(nc, 32119)

# B2 -- one plot each
p1 <- ggplot(nc)      + geom_sf(fill = "grey90", colour = "white") + theme_void()
p2 <- ggplot(nc_proj) + geom_sf(fill = "grey90", colour = "white") + theme_void()

p1
p2
# side by side, if you like:  library(patchwork);  p1 + p2
install.packages("patchwork")
library(patchwork)
p1 + p2

# B3 -- st_area() returns square METRES, as a units object
nc_proj <- nc_proj |>
  mutate(area_km2 = as.numeric(units::set_units(st_area(nc_proj), km^2)))

nc_proj |>
  st_drop_geometry() |>
  slice_max(area_km2, n = 5) |>
  select(NAME, area_km2)

# B4 -- now do the same starting from nc (degrees) and compare
nc_unProj <- nc |>
  mutate(area_km2 = as.numeric(units::set_units(st_area(nc), km^2)))

nc_unProj |>
  st_drop_geometry() |>
  slice_max(area_km2, n = 5) |>
  select(NAME, area_km2)
```

---

## Part C --- Choropleth maps (5 points)

This is the largest part of the exercise, and it is as much about judgement as
about code.

1. Add a column giving the **SIDS rate per 1,000 live births** for 1974--78.

2. Make a choropleth of the **raw counts** (`SID74`) and one of the **rate**,
   using `ggplot2`. Then write three or four sentences explaining why the two maps
   disagree, and which one you would put in front of a health department, and why.
The two maps disagree on which counties are struggling with SIDS because one uses a rate, and one uses a count. A count map does not control for population, so it could falsy signal to a health department that there's an issue in a heavily populated county when in actuality this county could have one of the lowest rates, and a rural county with a low overall count could have the highest rate and therefore go undetected. For this reason, I would show the map that uses rate74 instead of SID74 for the health department. 
   
   
3. Rebuild the rate map using **`tmap`**. Same information, other package.


4. Vary the **class intervals** on the rate map: produce it at least three ways
   --- for instance equal intervals, quantiles, and Jenks natural breaks. Two or
   three sentences on how much the apparent geography of risk changes. Name the
   classification you would defend, and say why.
The quantile rate map is not a good representation as the highest quanitle has such a large range as compared to other quanitiles due to outliers (2.78-9.55 is too large a range for the darkest color).For this reason, I favor the equal interval map as it shows clearly how severe each county is on an even scale, and can help highlight the worst counties easier. However, the equal interval map does possibly downplay some regions severity of SIDS death, so jenks may be a good option as well for highlighting the different problem regions. It does not show as many dark counties, but it does show more nuance then equal interval. 
   
5. Every map you submit must have a **legend with units**, a **title**, thin
   boundaries that do not overwhelm the fill, and a **perceptually uniform,
   colour-blind-safe palette**. State in one sentence which palette you chose and
   why, including whether dark means high or low.

ggplot(nc) +
  geom_sf(aes(fill = rate74), colour = "white", linewidth = 0.15) +
  scale_fill_distiller(palette = "Purples", direction = 1, name = "per 1,000 births") +
  labs(title = "Sudden Infant Deaths in North Carolina by County") +
  theme_void()
  
ggplot(nc) +
  geom_sf(aes(fill = rate74), colour = "white", linewidth = 0.15) +
  scale_fill_viridis_c(option = "magma", direction = -1, name = "per 1,000 births") +
  labs(title = "Sudden Infant Deaths in North Carolina by County (rate per 1,000)") +
  theme_void()
  
  These two are my favorites from each color package options (viridis and distiller) but I feel that the magma option on viridis is a better option as it uses multiple colors and can show a more nuanced color range, while still utilitizing lightness/darkness to make it more colorblind friendly. The purple brewer scale is nice, but I don't think it varies enough to show the worst counties. For both, I think it's best for dark to mean high rates as we associate darker tones with more danger. I also believe this is the precedent set by other health maps. 
  I chose a white border as already was written because it is not a color shown for the counties so it doens't confuse the reader or make the map look less professional. 
  
  
# I 





::: {.callout-tip}
## What earns full marks here
Not the prettiest map --- the map whose choices you can defend. A student who
picks quantiles and explains the trade-off in two clear sentences scores higher
than one who picks Jenks because it looked nicer.
:::

### Starting code

```{r}
#| label: c-starter

# C1 -- deaths per 1,000 live births
nc <- nc |>
  mutate(rate74 = 1000 * SID74 / BIR74 )

# C2 -- counts, then rates. Same code, one variable changed.
ggplot(nc) +
  geom_sf(aes(fill = SID74), colour = "white", linewidth = 0.15) +
  scale_fill_viridis_c(option = "magma", direction = -1, name = "count") +
  labs(title = "Sudden Infant Deaths in North Carolina by County (count)") +
  theme_void() +

ggplot(nc) +
  geom_sf(aes(fill = rate74), colour = "white", linewidth = 0.15) +
  scale_fill_viridis_c(option = "magma", direction = -1, name = "per 1,000 births") +
  labs(title = "Sudden Infant Deaths in North Carolina by County (rate per 1,000)") +
  theme_void()

# other palettes to try:
#   scale_fill_viridis_c(option = "viridis" / "plasma" / "cividis" / "mako")
#   scale_fill_distiller(palette = "YlOrRd", direction = 1)
#   RColorBrewer::display.brewer.all()   # shows every ColorBrewer option

# C3 -- the same map in tmap
tmap_mode("plot")                        # "view" makes it interactive

tm_shape(nc) +
  tm_polygons(fill = "rate74") +
  tm_title("Sudden Infant Deaths in North Carolina by county (rate per 1,000)")

# C4 -- change ONLY the classification, so the comparison is fair
tm_shape(nc) +
  tm_polygons(fill = "rate74",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5), 
              fill.legend = tm_legend(title = "SIDS Deaths per 1,000")) +
  tm_title("SIDS Deaths per 1,000, for each county in North Carolina, using Quantiles")    # does not give enough variability in the highest quantile, does a poor job of representing high-end outliers 

tm_shape(nc) +
  tm_polygons(fill = "rate74",
              fill.scale = tm_scale_intervals(style = "equal", n = 5), 
             fill.legend = tm_legend(title = "SIDS Deaths per 1,000")  ) +
  tm_title("SIDS Deaths per 1,000, for each county in North Carolina, using Equal interval")

tm_shape(nc) +
  tm_polygons(fill = "rate74",
              fill.scale = tm_scale_intervals(style = "jenks", n = 5), 
             fill.legend = tm_legend(title = "SIDS Deaths per 1,000")  ) +
  tm_title("SIDS Deaths per 1,000, for each county in North Carolina, using Jenks")

tm_shape(nc) +
  tm_polygons(fill = "rate74",
              fill.scale = tm_scale_intervals(style = "sd", n = 5), 
             fill.legend = tm_legend(title = "SIDS Deaths per 1,000")  ) +
  tm_title("SIDS Deaths per 1,000, for each county in North Carolina, using sd")


# then repeat with style = "equal" and style = "jenks"
# other options: "pretty", "fisher"

# question 5c 
ggplot(nc) +
  geom_sf(aes(fill = rate74), colour = "white", linewidth = 0.15) +
  scale_fill_distiller(palette = "Purples", direction = 1, name = "per 1,000 births") +
  labs(title = "Sudden Infant Deaths in North Carolina by County") +
  theme_void()

ggplot(nc) +
  geom_sf(aes(fill = rate74), colour = "white", linewidth = 0.15) +
  scale_fill_viridis_c(option = "magma", direction = -1, name = "per 1,000 births") +
  labs(title = "Sudden Infant Deaths in North Carolina by County (rate per 1,000)") +
  theme_void()
  



```

::: {.callout-warning}
## A note on `tmap` versions
`tmap` changed its argument names in version 4. If an example you find online
uses `tm_polygons(col = ...)` or `style = ...` directly and yours errors, check
`packageVersion("tmap")` and read the help page rather than guessing. The code
above is written for version 4.
:::

---

## Part D --- A point layer (2 points)

1. Look up the approximate longitude and latitude of **four cities in North
   Carolina** and build a small table holding the name and the two coordinates.
   Cite where you got the coordinates.
   
   
2. Turn that table into an `sf` object with the correct CRS. State which EPSG
   code you used and why.
   
I used EPSG 4326 because the table uses longitude and latitude coordinates and from what I've learned so far, that is the correct CRS for this geographic coordinate system. 
   
   
3. Plot the four cities **on top of** your rate choropleth, labelled legibly.


4. One sentence: what would go wrong if you built the `sf` object with the
   coordinate columns in the other order?

If we switched the lon and lat coordinates, the points would be placed in the wrong location. 

### Starting code

```{r}
#| label: d-starter

# D1 -- North Carolina longitudes are NEGATIVE (west of Greenwich)
cities <- tibble(
  name = c("Winston-Salem", "Asheville", "Greensboro", "Fayetteville"),
  lon  = c(-80.244, -82.554, -79.766, -78.878),
  lat  = c(36.099, 35.601, 36.045, 35.053)
)
cities

# D2 -- coords takes the column names in x, y order: longitude FIRST
cities_sf <- cities |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

# D3 -- layers draw in order, so points come after polygons
ggplot() +
  geom_sf(data = nc, aes(fill = rate74), colour = "white", linewidth = 0.15) +
  geom_sf(data = cities_sf, size = 1, colour = "black") +
  geom_sf_text(data = cities_sf, aes(label = name), size = 3, nudge_y = 0.15, nudge_y = 0.2, color = "black", fontface = "bold") +
  scale_fill_viridis_c(option = "magma", direction = -1, name = "SIDS deaths per 1,000 births") +
  labs(title = "Sudden Infant Deaths in North Carolina by County" )+
  theme_void()

# check your cities landed in North Carolina and not in the ocean
```

---

## Part E --- Reproducibility and AI use (1 point)

1. Your document renders from a clean session, with every package loaded
   explicitly. *(Check this last, after you have stopped editing.)*
   
   
2. End the document with an **AI use note**: which tools you used, what for, and
   at which stage --- planning, coding, debugging, interpreting, or writing. Two
   or three sentences.
   
   I used AI to explain the concept of a CRS/EPSG code and questions to better understand some of the course concepts. I also used it for debugging a couple times but did not copy/paste any code, I just used it to better understand. Both errors were due to syntax issues and not substantial. 

Remember the course policy: AI use is unrestricted and costs you nothing, but you
must be able to explain every line you hand in. Disclosure is the only
requirement.

### Starting code

```{r}
#| label: e-starter

sessionInfo()
```

---

## Rubric

| Part | What is assessed | Points |
|---|---|---|
| **A** | Correct inspection of the object; clear, accurate answers | 2 |
| **B** | Correct transformation and areas; comparison explained | 2 |
| **C** | Both packages working; counts-vs-rates argument; class intervals compared and a defended choice; cartographic standards met | 5 |
| **D** | Point layer built with the right CRS and correctly overlaid | 2 |
| **E** | Renders from a clean session; AI use note present | 1 |
| | **Total** | **12** |

Writing counts for roughly 10% of the mark within each part: the prose should be
in complete sentences and free of obvious errors. Every part asks for
interpretation as well as code --- a submission that is all code and no sentences
cannot score above about 60%.

---

## Getting help

- Office hours, or email me --- with the code and the exact error message, not a
  screenshot of the whole console.
- The `sf` documentation is genuinely good: `vignette("sf1", package = "sf")`.
- Moraga Chapter 2 covers most of Parts A--C.
- If you are stuck for more than about thirty minutes on something that feels
  mechanical, that is a signal to ask rather than to persevere.